---
title: "Analyses Cerastium"
output:
  pdf_document: 
    toc: yes
editor_options: 
  chunk_output_type: inline
---
\newpage
```{r Load libraries, include=FALSE}
library(ggplot2)
library(ggthemes)
library(car)
library(grid)
library(gridExtra)
library(papeR)
library(knitr)
library(lme4)
library(lmerTest)
library(lubridate)
library(dplyr)
library(agricolae)
library(effects)
```

```{r Define ggplot theme, include=FALSE}
my_theme <- function(){
  theme_base()+theme(plot.background=element_rect(fill="white", colour=NA))+
  theme(legend.position="none")+theme(text=element_text(family="serif"))
}
```

```{r Read data, include=FALSE}
#Read data files 2015 (clean files after 1_data_prep)
data_f_2015<-read.table("C:/Users/User/Dropbox/SU/Projects/cerastium/data/clean/cerastium_field_clean.txt",
                        header=T,sep="\t",dec=".")
data_f_2017<-read.table("C:/Users/User/Dropbox/SU/Projects/cerastium/data/clean/cerastium_2017_FFD_temp_fit.txt",
                        header=T,sep="\t",dec=".")
data_cg<-read.table("C:/Users/User/Dropbox/SU/Projects/cerastium/data/clean/cerastium_cgarden_clean.txt",
                    header=T,sep="\t",dec=".")

data_f_2015$n_flowered<-with(data_f_2015,ifelse(stage==5,0,1)) #Probability of NOT having flowered
data_cg$first_fl<-as.Date(as.character(data_cg$first_fl))
data_cg$last_fl<-as.Date(as.character(data_cg$last_fl))
data_cg$first_bud<-as.Date(as.character(data_cg$first_bud))
data_cg$last_bud<-as.Date(as.character(data_cg$last_bud))
data_cg$peak_fl<-as.Date(as.character(data_cg$peak_fl))
data_cg$patch_new<-as.factor(paste(data_cg$stream,data_cg$patch,sep="_"))
data_cg$first_fl_j<-yday(data_cg$first_fl)
data_cg$mother_pl_id_new<-as.factor(data_cg$mother_pl_id_new)
```


```{r recode streams and plots, include=FALSE}
data_f_2015$stream_new<-as.factor(with(data_f_2015,ifelse(stream=="17","1",
                        ifelse(stream=="Main","2",ifelse(stream=="Parking","3","4")))))
data_cg$stream_new<-as.factor(with(data_cg,ifelse(stream=="17","1",
                    ifelse(stream=="Main","2",ifelse(stream=="Park","3","4")))))
data_f_2017$plot_new<-factor(with(data_f_2017,ifelse(plot=="H01","1",ifelse(plot=="H02","2",
                      ifelse(plot=="H03","3",ifelse(plot=="H04","4",ifelse(plot=="H05","5",
                      ifelse(plot=="H08","6",ifelse(plot=="H09","7",ifelse(plot=="H10","8",
                      ifelse(plot=="H13","9",ifelse(plot=="HC1","10",ifelse(plot=="HC2a"|plot=="HC2b","11",
                      ifelse(plot=="HC4","12",ifelse(plot=="HC5","13","14")))))))))))))),
                      levels=c("1","2","3","4","5","6","7","8","9","10","11","12","13","14"),ordered=T)
```



# Field data: Distributions

```{r Distributions, echo=FALSE, fig.height=3.5, fig.width=10, warning=FALSE}
# Make plots
distrs_2015<-grid.arrange(
  ggplot(data_f_2015,aes(stage))+geom_histogram(stat="count",colour="black", fill="grey")+xlab("Phenological\nstage")+ylab("Count")+my_theme()+ggtitle("2015"),
  ggplot(data_f_2015,aes(n_flowered))+geom_histogram(stat="count",colour="black",fill="grey")+xlab("Probability\nof not flowering")+ylab("Count")+my_theme()+ggtitle(""),
  ggplot(data_f_2015,aes(temp))+geom_histogram(binwidth=5,colour="black", fill="grey")+xlab("Soil temperature\n(ºC)")+ylab("Count")+my_theme()+ggtitle(""),
ncol=3)

blank<-grid.rect(gp=gpar(col="white"))

distrs_2017<-grid.arrange(
  ggplot(data_f_2017,aes(yday(FFD)))+geom_histogram(binwidth=5,colour="black",fill="grey")+xlab("FFD\n")+ylab("Count")+my_theme()+ggtitle("2017"),
  blank,
  ggplot(data_f_2017,aes(temp1))+geom_histogram(binwidth=5,colour="black", fill="grey")+xlab("Soil temperature\n(ºC)")+ylab("Count")+my_theme()+ggtitle(""),
ncol=3)

# Save plots
ggsave(filename="C:/Users/User/Dropbox/SU/Projects/cerastium/results/figures/distrs_2015.tiff",plot=distrs_2015,device="tiff",width=30,height=10,units="cm",dpi=600)
ggsave(filename="C:/Users/User/Dropbox/SU/Projects/cerastium/results/figures/distrs_2017.tiff",plot=distrs_2017,device="tiff",width=30,height=10,units="cm",dpi=600)
```

# Field data: Differences among streams

## 2015

```{r Plots differences among streams 2015, echo=FALSE, fig.height=3.5, fig.width=10,warning=FALSE}
p1<-ggplot(data_f_2015,aes(x=stream_new,y=stage))+geom_boxplot()+xlab("Strean")+ylab("Phenological stage")+my_theme()
p2<-ggplot(data_f_2015,aes(x=stream_new,fill=as.factor(n_flowered)))+geom_bar(position="dodge")+xlab("Stream")+ylab("Count")+my_theme()+theme(legend.position=c(0.8, 0.82),legend.title=element_blank())
p3<-ggplot(data_f_2015,aes(x=stream_new,y=temp))+geom_boxplot()+xlab("Stream")+ylab("Soil temperature (ºC)")+my_theme()
grid.arrange(p1,p2,p3,ncol=3)
ggsave(filename="C:/Users/User/Dropbox/SU/Projects/cerastium/results/figures/diffs_2015.tiff",
       plot=grid.arrange(p1,p2,p3,ncol=3),device="tiff",width=30,height=10,units="cm",dpi=600) 
```

Differences among streams in phenological stage: Anova and Tukey HSD

```{r stage~stream 2015, fig.height=2.2, fig.width=3,message=FALSE}
kable(prettify(with(data_f_2015,Anova(lm(stage~stream_new)))))
kable(prettify(as.data.frame(with(data_f_2015,TukeyHSD(aov(lm(stage~stream_new))))$stream_new)))
```

Differences among streams in proability of flowering: Anova and Tukey HSD

```{r n_flowered~stream 2015, fig.height=2.2, fig.width=3, warning=FALSE,message=FALSE}
kable(prettify(with(data_f_2015,Anova(glm(n_flowered~stream_new,family="binomial")))))
kable(prettify(as.data.frame(with(data_f_2015,TukeyHSD(aov(as.numeric(n_flowered)~stream_new)))$stream_new)))
```

Differences among streams in temperature: Anova and Tukey HSD

```{r temp~stream 2015, fig.height=3, fig.width=4, warning=FALSE,message=FALSE}
kable(prettify(with(data_f_2015,Anova(lm(temp~stream_new)))))
Tukey_temp_2015<-with(data_f_2015,TukeyHSD(aov(lm(temp~stream_new))))
kable(prettify(as.data.frame(Tukey_temp_2015$stream_new)))
```

## 2017

```{r Plots differences among plots 2017, echo=FALSE, fig.height=3.5, fig.width=10,warning=FALSE,message=FALSE}
p4<-ggplot(data_f_2017,aes(x=plot_new,y=yday(FFD)))+geom_boxplot()+xlab("Plot")+ylab("FFD")+my_theme()
p5<-ggplot(data_f_2017,aes(x=plot_new,y=temp1))+geom_boxplot()+xlab("Plot")+ylab("Soil temperature (ºC)")+my_theme()
grid.arrange(p4,p5,ncol=2)
ggsave(filename="C:/Users/User/Dropbox/SU/Projects/cerastium/results/figures/diffs_2017.tiff",
       plot=grid.arrange(p4,p5,ncol=2),device="tiff",width=30,height=10,units="cm",dpi=600) 
```

Differences among plots in FFD: Anova and Tukey HSD

```{r FFD~plot 2017, fig.height=2.2, fig.width=3, warning=FALSE,message=FALSE}
kable(prettify(with(data_f_2017,Anova(lm(yday(FFD)~plot_new)))))
#kable(prettify(as.data.frame(with(data_f_2017,TukeyHSD(aov(yday(FFD)~plot)))$plot)))
```

Differences among plots in temperature: Anova and Tukey HSD

```{r temp~plot 2017, fig.height=3, fig.width=4, warning=FALSE,message=FALSE}
kable(prettify(with(data_f_2017,Anova(lm(temp1~plot_new)))))
#kable(prettify(as.data.frame(with(data_f_2017,TukeyHSD(aov(lm(temp1~plot))))$plot)))
```

Differences among plots in temperature with results from Tukey HSD: 2015 and 2017

```{r boxplots paper, echo=FALSE}
data_f_2015$stream_ord<-with(data_f_2015,ifelse(stream_new==1,1,
            ifelse(stream_new==3,2,
                   ifelse(stream_new==2,3,4))))
data_f_2017$plot_ord<-with(data_f_2017,ifelse(plot_new==12,1,
            ifelse(plot_new==10,2,
              ifelse(plot_new==14,3,
                ifelse(plot_new==13,4,
                   ifelse(plot_new==11,5,
                      ifelse(plot_new==7,6,
                         ifelse(plot_new==3,7,
                            ifelse(plot_new==6,8,
                               ifelse(plot_new==2,9,
                                  ifelse(plot_new==8,10,
                                     ifelse(plot_new==5,11,
                                        ifelse(plot_new==9,12,
                                            ifelse(plot_new==4,13,14))))))))))))))

hsd_15=HSD.test(aov(temp~stream_ord,data=data_f_2015), "stream_ord", group=T)
data_f_2015.summarized = data_f_2015 %>% group_by(stream_ord) %>% summarize(Max.temp=max(temp))

hsd_17=HSD.test(aov(temp1~plot_ord,data=data_f_2017), "plot_ord", group=T)
data_f_2017.summarized = data_f_2017 %>% group_by(plot_ord) %>% summarize(Max.temp=max(temp1))

boxplots<-grid.arrange(
  ggplot(data_f_2015,aes(x=reorder(stream_ord,temp,FUN=mean),y=temp))+geom_boxplot(outlier.shape = NA)+
    xlab("Plot")+ylab("Soil temperature (ºC)")+scale_y_continuous(limits=c(0,50))+
    geom_text(data=data_f_2015.summarized,aes(x=stream_ord,y=0.6+Max.temp,label=hsd_15$groups$groups),vjust=0)+
    my_theme()+ggtitle("A) 2015")+stat_summary(fun.y=mean, geom="point",size=1.5),
  ggplot(data_f_2017,aes(x=reorder(plot_ord,temp1,FUN=mean),y=temp1))+geom_boxplot(outlier.shape = NA)+
    xlab("Plot")+ylab(NULL)+scale_y_continuous(limits=c(0,50))+
    geom_text(data=data_f_2017.summarized,aes(x=plot_ord,y=0.6+Max.temp,label=hsd_17$groups$groups),vjust=0)+
    my_theme()+ggtitle("B) 2017")+stat_summary(fun.y=mean, geom="point",size=1.5),
  ncol=2,widths=c(1.2,3))
ggsave(filename="C:/Users/User/Dropbox/SU/Projects/cerastium/results/figures/diffs_temp.tiff",
       plot=boxplots,device="tiff",width=30,height=10,units="cm",dpi=600)
```


# Hypothesis 1: Local soil warming leads to an earlier phenology

```{r phenology~temp plot, echo=FALSE,fig.height=4.5, fig.width=9,warning=FALSE}
p6<-ggplot(data_f_2015,aes(x=temp,y=n_flowered))+geom_point()+
  geom_smooth(method = "glm", method.args=c(family="binomial"),color="black",size=1.5)+
  my_theme()+xlab("Soil temperature")+ylab("Probability of not having flowered\nat the day of recording")+
  # theme(legend.position=c(0.83,0.3))+
  ggtitle("A) 2015")+scale_x_continuous(limits=c(5,35))
p7<-ggplot(data_f_2017,aes(x=temp1,y=yday(FFD)))+geom_point()+
  geom_smooth(method = "lm",color="black",size=1.5)+
  my_theme()+xlab("Soil temperature")+ylab("FFD (Julian date)")+
  # theme(legend.position="right")+
  ggtitle("B) 2017")
grid.arrange(p6,p7,ncol=2)
ggsave(filename="C:/Users/User/Dropbox/SU/Projects/cerastium/results/figures/phen_temp.tiff",
       plot=grid.arrange(p6,p7,ncol=2),device="tiff",width=22,height=10,units="cm",dpi=600) 
```

## 2015

### Logistic regression (relationship among probability of flowering and soil temperature) with pooled data 

```{r n_flowered~temp pooled data, echo=TRUE, message=FALSE, warning=FALSE}
kable(prettify(summary(glm(n_flowered~temp,data=data_f_2015,family="binomial"))))
```

There is a strong overall relatiohsnip among probability of flowering and soil temperature.

### Logistic regression for each stream

```{r n_flowered~temp each stream, echo=TRUE, message=FALSE, warning=FALSE}
kable(prettify(summary(glm(n_flowered~temp,data=subset(data_f_2015,stream_new=="1"),family="binomial"))))
kable(prettify(summary(glm(n_flowered~temp,data=subset(data_f_2015,stream_new=="2"),family="binomial"))))
kable(prettify(summary(glm(n_flowered~temp,data=subset(data_f_2015,stream_new=="3"),family="binomial"))))
kable(prettify(summary(glm(n_flowered~temp,data=subset(data_f_2015,stream_new=="4"),family="binomial"))))
```

Within each stream, there are no significant relationships.

### Logistic regression including temperature and stream

```{r n_flowered~temp+stream, message=FALSE}
model1<-glmer(n_flowered~temp+(1|stream_new),family="binomial",data_f_2015)
kable(prettify(summary(model1)))
anova(model1,glm(n_flowered~temp,data=data_f_2015,family="binomial"))
```

Random effect of plot is significant according to likelihood ratio test.
We cannot really separate the effects of streams from the effects of temperature, as streams have very different ranges of temperatures, i.e. effects of streams are likely to be mostly effects of temperatures. This means that differences in temperature at larger scales (among streams) are more important than differences at small scales (within streams). 

## 2017

### Linear regression (relationship among FFD and soil temperature) with pooled data 

```{r FFD~temp pooled data, echo=TRUE, message=FALSE, warning=FALSE}
kable(prettify(summary(lm(yday(FFD)~temp1,data=data_f_2017))))
```

There is a strong overall relatiohsnip among FFD and soil temperature.

### Linear regression including temperature and plot

```{r FFD~temp+plot, message=FALSE}
model2<-lmer(yday(FFD)~temp1+(1|plot_new),data_f_2017,REML=F)
kable(prettify(summary(model2)))
anova(model2,lm(yday(FFD)~temp1,data=data_f_2017))
```

Random effect of plot is significant according to likelihood ratio test.
We cannot really separate the effects of plots from the effects of temperature, as plots have very different ranges of temperatures, i.e. effects of plots are likely to be mostly effects of temperatures. This means that differences in temperature at larger scales (among plots) are more important than differences at small scales (within plots). 

# Hypothesis 2: Phenotypic selection for early flowering is stronger in colder sites with shorter growing seasons

Select data for selection analyses

```{r subset data 2017 selection}
# Select plants with no NA values
data_f_2017_sel<-subset(data_f_2017,!is.na(FFD)&!is.na(n_seeds)&!is.na(nfl)) 
nrow(data_f_2017)     # N data 2017
nrow(data_f_2017_sel) # N data used for selection analyses 2017 
#Standardized traits
data_f_2017_sel$FFD_std<-(yday(data_f_2017_sel$FFD)-
                            mean(yday(data_f_2017_sel$FFD)))/sd(yday(data_f_2017_sel$FFD))
data_f_2017_sel$nfl_std<-(data_f_2017_sel$nfl-
                            mean(data_f_2017_sel$nfl))/sd(data_f_2017_sel$nfl)
#Relative fitness
data_f_2017_sel$n_seeds_rel<-data_f_2017_sel$n_seeds/mean(data_f_2017_sel$n_seeds)
```

Selection models

```{r selection models}
model5<-lm(n_seeds_rel~FFD_std*temp1+nfl_std,data=data_f_2017_sel)
kable(prettify(summary(model5))) # Selection for FFD does not depend on temperature
model6<-lm(n_seeds_rel~FFD_std+temp1+nfl_std,data=data_f_2017_sel) 
kable(prettify(summary(model6))) # Interaction removed --> selection for early flowering
model7<-lm(n_seeds_rel~FFD_std*plot_new,data=data_f_2017_sel)
kable(prettify(Anova(model7))) # No differences in selection gradients for FFD among plots
```

There is no interactive effect of phenology and soil temperature on fitness, i.e. selection on phenology does not depend on temperature. 
Fitness increases with early flowering, independent of temperature, i.e. there is overall selection for early flowering.
This could be the result of environmental covariation, i.e. that both early flowering and fitness are correlated with favorable environmental conditions.

```{r Selection models with log fitness and plot as random}
summary(lmer(log(n_seeds+1)~FFD_std*temp1+nfl_std+(1|plot_new),data=data_f_2017_sel))
```


```{r Fitness vs. FFD, echo=FALSE}
ggplot(data_f_2017_sel,aes(x=yday(FFD),y=log(n_seeds+1)))+geom_point()+
  geom_smooth(method = "lm", color="black",size=1.5)+
  my_theme()+xlab("FFD (Julian date)")+ylab("Number of seeds (log)")
ggsave(filename="C:/Users/User/Dropbox/SU/Projects/cerastium/results/figures/fitness_FFD.tiff",
       device="tiff",width=12,height=12,units="cm",dpi=600) 
```

```{r Fitness vs. FFD and interaction with temp, echo=FALSE}
data_f_2017_sel$FFD_julian<-yday(data_f_2017_sel$FFD)
sel_model<-lmer(log(n_seeds+1)~FFD_julian+temp1+FFD_julian:temp1+nfl_std+(1|plot_new),data=data_f_2017_sel)
interaction1<-data.frame(effect(term="FFD_julian:temp1", mod=sel_model,
                                xlevels=list(FFD_julian=seq(156,211,1), temp1=seq(4,33,1))))
ggplot(interaction1, aes(FFD_julian,fit, group = as.factor(temp1)))+
  geom_smooth(method=loess,se=F,size=0.3,aes(FFD_julian,fit,color=temp1))+
  xlab("FFD (Julian date")+ylab("Number of seeds (log)")+theme_base()+
  scale_color_gradient(low = "blue", high = "red")+
  theme(legend.position="none")+labs(colour="Soil temperature")+
  theme(text=element_text(family="serif"))+theme(plot.background=element_rect(fill="white", colour=NA))
ggsave(filename="C:/Users/User/Dropbox/SU/Projects/cerastium/results/figures/fitness_FFD_temp.tiff",
       device="tiff",width=12,height=12,units="cm",dpi=600) 
```


# Hypothesis 3: As a consequence of selection, differences in phenology in a common environment are related to soil temperature at the plant origin in a counter-gradient fashion, with plants originating from colder sites flowering earlier

Histogram of Julian date of first flowering in the common garden (2017)

```{r distr first_fl_j, echo=FALSE,warning=FALSE,message=FALSE,fig.height=3.5, fig.width=4.5}
ggplot(data_cg,aes(first_fl_j))+geom_histogram(colour="black", fill="grey")+my_theme()+xlab("Julian date of first flowering")+ylab("Count")
ggsave(filename="C:/Users/User/Dropbox/SU/Projects/cerastium/results/figures/distr_FFD_cg.tiff",
       device="tiff",width=12,height=10,units="cm",dpi=600) 
```

## Effect of mother plant (and stream) on the flowering phenology in the common garden

Both mother plant and stream are treated as random effects. 

```{r first_fl_j~stream+mother, warning=FALSE, message=FALSE}
# This takes into account that mother is nested within stream
model8<-lmer(first_fl_j~(1|stream_new/mother_pl_id_new), REML=T,data=data_cg)
summary(model8)
# Likelihood ratio test for testing the signigicance of random effects
# Compares a model with a given random effect to the same model without the random effect
ranova(model8) # Mother is significant
```

There is an effect of both mother plant and stream of origin on flowering phenology (first flowering day) in the common garden: there are differences among streams, and differences among mothers within streams.

## Are differences among mother plants on the flowering phenology in the common garden related to soil temperature at the site of origin?

We expect plants to respond to warming in spring in a counter-gradient fashion, i.e. plants growing on warmer soils will flower early in the field but later in the common garden (requiring more warm days to start development) than plants from cold soils. Plants growing on colder soils will flower later in the field, but earlier in the common garden, because they can start developing at lower temperatures (they are adapted to colder conditions and they have evolved to compensate for the later and shorter growing season). They have the genetic capacity to develop at lower temperatures, probably because they have been selected for rapid development in environments with short growing seasons.

```{r first_fl_j~temp plot, echo=FALSE,fig.height=3.5, fig.width=4.5,warning=FALSE}
ggplot(data_cg,aes(x=temp_ori,y=first_fl_j))+geom_point()+
  geom_smooth(method = "lm", color = "black",size=1.5)+ 
  my_theme()+xlab("Soil temperature at origin")+ylab("FFD (Julian date)\nin the common garden")
ggsave(filename="C:/Users/User/Dropbox/SU/Projects/cerastium/results/figures/FFD_tempori.tiff",
       device="tiff",width=12,height=10,units="cm",dpi=600) 
```

### LM with pooled data

```{r first_fl_j~temp_ori pooled data, echo=TRUE, message=FALSE, warning=FALSE}
kable(prettify(summary(lm(first_fl_j~temp_ori,data=data_cg))))
```

There is a strong overall relatiohsnip among FFD in the common garden and soil temperature at the origin, that goes in the expected direction (i.e. opposite to the relationship observed in field conditions).

### LMM for each stream

```{r first_fl_j~temp_ori each stream, echo=TRUE, message=FALSE, warning=FALSE}
kable(prettify(summary(lmer(first_fl_j~temp_ori+(1|mother_pl_id_new),data=subset(data_cg,stream_new=="1")))))
kable(prettify(summary(lmer(first_fl_j~temp_ori+(1|mother_pl_id_new),data=subset(data_cg,stream_new=="2")))))
kable(prettify(summary(lmer(first_fl_j~temp_ori+(1|mother_pl_id_new),data=subset(data_cg,stream_new=="3")))))
kable(prettify(summary(lmer(first_fl_j~temp_ori+(1|mother_pl_id_new),data=subset(data_cg,stream_new=="4")))))

```

Within each stream, there are no significant relationships.

### LMM including temperature, stream and mother

```{r first_fl_j~temp_ori+stream+mother, warning=FALSE, message=FALSE}
# This takes into account that mother is nested within stream
model9<-lmer(first_fl_j~temp_ori+(1|stream_new/mother_pl_id_new),data = data_cg)
kable(prettify(summary(model9))) # Temperature is significant
ranova(model9) #Stream and mother are significant
```

Differences among mother plants in phenology are related to soil temperature at origin. Plants respond to warming a counter-gradient fashion: plants from warmer soils flower later in the common garden than plants from cold soils. 

# Phenotypic and genotypic responses

```{r Phenotypic and genotypic, echo=FALSE}
ggplot()+
geom_smooth(data=data_f_2017,aes(x=temp1,y=yday(FFD)),method = "lm",color="black",size=1.5)+
  my_theme()+xlab("Soil temperature")+ylab(NULL)+
    theme(axis.text.y=element_blank())
ggsave(filename="C:/Users/User/Dropbox/SU/Projects/cerastium/results/figures/phen_temp_17.tiff",
       device="tiff",width=8,height=8,units="cm",dpi=600) 
ggplot(data_cg,aes(x=temp_ori,y=first_fl_j))+
  geom_smooth(method = "lm", color = "black",size=1.5)+ 
  my_theme()+xlab("Soil temperature at origin")+ylab(NULL)+
  theme(axis.text.y=element_blank())
ggsave(filename="C:/Users/User/Dropbox/SU/Projects/cerastium/results/figures/phen_temp_cg.tiff",
       device="tiff",width=8,height=8,units="cm",dpi=600) 
```

